---
title: "Case Study 2"
author: "First Name, Last Name"
date: "2025-04-02"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
*The  .Rmd* **and** *.html (or .pdf) should be uploaded in TUWEL by the deadline. Refrain from using explanatory comments in the R code chunks but write them as text instead. Points will be deducted if the submitted file is not in a decent form.*

**DISCLAIMER**: In case students did not contribute equally, include a disclaimer stating what each student's contribution was.


The CIA World Factbook provides intelligence on various aspects of 266 world entities, including history, people, government, economy, energy, geography, environment, communications, transportation, military, terrorism, and transnational issues. This case study involves analyzing world data from 2020, focusing on:

- **Education Expenditure (% of GDP)**
- **Youth Unemployment Rate (15-24 years)**
- **Net Migration Rate** (difference between the number of people entering and leaving a country per 1,000 persons)

The data was sourced from the [CIA World Factbook Archives](https://www.cia.gov/the-world-factbook/about/archives/). You are required to use `dplyr` for data manipulation, while any package can be used for importing data.

# Tasks:

## a. Data Import and Cleaning

Load the following datasets from TUWEL and ensure that missing values are handled correctly and column names are clear. Each dataset should ultimately contain only two columns: **country** and the respective variable. Note that some data sets also contain information on the year when the value was last updated.

```{r setup-import}
# Load the tidyverse package collection
library(tidyverse)
library(dbplyr)

# Set path to data folder
data_path <- "data"
# data_path <- "src/Case_Study_2/data" anscheidnend für knit
```
  
  * `rawdata_369.txt` which contains the (estimated)  public expenditure on education as a percent of GDP. *Pay attention! The delimiter is 2 or more white spaces (one space would not work as it would separate country names which contain a space); you have to skip the first two lines*. 
  
```{r import-edu-clean}
# Read lines from file, skip first two lines (title and separator)
edu_lines <- readLines(file.path(data_path, "rawdata_369.txt"))[-c(1, 2)]

# Convert to tibble with fixed-width fields
edu_exp <- tibble(raw = edu_lines) %>%
  separate(raw, into = c("index", "country", "education_expenditure", "year"), 
           sep = "\\s{2,}", extra = "merge", fill = "right") %>%
  select(country, education_expenditure, year) %>%
  mutate(
    country = str_to_lower(country),
    country = str_remove_all(country, "–|−|-"),       # remove various dashes
    country = str_squish(country),                    # normalize whitespace
    education_expenditure = as.numeric(education_expenditure),
    year = as.integer(year)
  )
```

  * `rawdata_373.csv` which contains the (estimated) youth unemployment rate (15-24) per country
  
```{r import-unemployment}
# Read youth unemployment data from CSV

youth_unemp <- read_csv(
  file.path(data_path, "rawdata_373.csv"),
  show_col_types = FALSE
) %>%
  select(
    country = 1,
    youth_unemployment = 2
  ) %>%
  mutate(
    country = str_to_lower(country),  # make country names lowercase
    country = str_replace_all(country, "timor\\-?leste", "timorleste")  # unify country name, because of odd duplicates
  )
```
  
  * `rawdata_347.txt`  which contains (estimated) net migration rate per country.
  
```{r import-migration}
# Read lines from file, skip first two lines (header and separator)
mig_lines <- readLines(file.path(data_path, "rawdata_347.txt"))[-c(1, 2)]

# Convert to tibble and split into components
net_mig <- tibble(raw = mig_lines) %>%
  separate(raw, into = c("index", "country", "net_migration", "year_raw"), 
           sep = "\\s{2,}", extra = "merge", fill = "right") %>%
  select(country, net_migration, year_raw) %>%
  mutate(
    country = str_to_lower(country),                 # make country names lowercase
    country = str_remove_all(country, "–|−|-"),      # remove dashes
    country = str_squish(country),                   # normalize whitespace
    net_migration = as.numeric(net_migration),        # convert to numeric
    year = str_extract(year_raw, "\\d{4}"),     # extract the 4-digit year
    year = as.integer(year)
  ) %>%
  select(-year_raw)
```


## b. Merging Raw Data

Merge the datasets using `dplyr` on a unique key and retain the union of all observations.

```{r merge-data}
# Rename year columns before joining
edu_exp_clean <- edu_exp %>%
  rename(edu_year = year)

net_mig_clean <- net_mig %>%
  rename(mig_year = year)

# Merge all datasets on 'country'
merged_data <- youth_unemp %>%
  full_join(edu_exp_clean, by = "country") %>%
  full_join(net_mig_clean, by = "country") %>%
  arrange(country)
```

- What key are you using for merging? -> Country
- Return the dimensions of the merged dataset.

```{r preview-merged-dim}
# Check structure and dimensions of the merged dataset
dim(merged_data)
```

```{r preview-merged-table}
merged_data
```

## c. Enriching Data with Income Classification

Obtain country income classification (low, lower-middle, upper-middle, high) from the [World Bank](https://datahelpdesk.worldbank.org/knowledgebase/articles/906519) and merge it with the dataset.

- Identify common variables between datasets. Can they be used for merging? Why or why not?
  The only common variable between our existing merged_data and the data from the world bank,
  is the country name. Although the latter doesn't use lowercase names for the countries.
  
```{r}
library(readxl)
world_bank_income <- read_excel(file.path(data_path, "CLASS.xlsx"),
                    sheet = "List of economies",
                    col_names = TRUE) %>%
                    select(Code, Economy, `Income group`)

world_bank_income <- world_bank_income %>%
  mutate(Economy = tolower(Economy),
         Economy = str_trim(Economy))

merged_data_income <- merged_data %>%
  left_join(world_bank_income, by = c("country" = "Economy"))

data_without_income <- merged_data_income %>% 
  filter(is.na(`Income group`))

nrow(data_without_income)
```
  As seen above merging World Bank data with our existing data, results in 43 countries
  not receiving the corresponding data. Does that doesn't mean those countries
  don't exist in the World Bank data, but that they carry a different name. 

- Since ISO codes are standardized, download and use the [CIA country data codes](https://www.cia.gov/the-world-factbook/references/country-data-codes/) for merging. Make sure you are not losing any of the countries in your original data set when merging.

```{r}
country_codes <- read.csv(file.path(data_path, "Country_Data_Codes.csv"), stringsAsFactors = FALSE) %>%
  select(GENC, Name)

clean_income_data <- world_bank_income %>%
  left_join(country_codes, by = c("Code" = "GENC")) %>%
  mutate(Economy = ifelse(!is.na(Name), Name, Economy)) %>%  
  select(-Name)

clean_income_data <- clean_income_data %>%
  mutate(Economy = tolower(Economy),
         Economy = str_trim(Economy))

merged_data_income_clean <- merged_data %>%
  left_join(clean_income_data, by = c("country" = "Economy"))

data_without_income_clean <- merged_data_income_clean %>% 
  filter(is.na(`Income group`))

nrow(data_without_income_clean)
```
This time we receive a much lower result of countries, which don`t have a 
corresponding value. 


## d.  Adding Geographical Information

Introduce continent and subcontinent (or region) data for each country.

- Find and download an appropriate online resource.
  
  The United Nations Statistics Division offers a data set fitting the criteria
```{r}
un_stat <- read.csv(file.path(data_path, "UNSD-Methodology.csv"), sep = ";", stringsAsFactors = FALSE) %>%
  rename(
    Continent = `Region.Name`,
    Subcontinent = `Sub.region.Name`,
    iso3 = `ISO.alpha3.Code`
  ) %>%
  select(Continent, Subcontinent, iso3)
```
  
- Merge this information into the dataset, naming the final dataset `df_vars`. Make sure you are not losing any of the countries in your original data set when merging.

```{r}
df_vars <- merged_data_income_clean %>%
  left_join(un_stat, by = c("Code" = "iso3"))

summary(df_vars)

```



## e. Data Tidiness and Summary Statistics

- Evaluate the tidiness of `df_vars`
• Column headers are values, not variable names.✅
• Multiple types of observational units are stored in the same table.✅
• A single observational unit is stored in multiple tables.✅
• Variables are stored in both rows and columns.✅
(observational units, variables, fixed vs. measured variables).
Make adjustments to tidy the data, if necessary.
```{r}
library("tidyverse")
library(dplyr)
str(df_vars)

```
• Multiple variables are stored in one column.✅
```{r}
for (col in names(df_vars)) {
  cat("Unique values in column", col, ":\n")
  print(unique(df_vars[[col]]))
  cat("\n")
}
```
- Create a frequency table for the income status variable and briefly interpret the results.
```{r}
table(df_vars$`Income group`)
```



- Analyze the distribution of income status across continents by computing absolute and relative frequencies. Comment on the findings.
```{r}
for (cont in unique(df_vars$Continent)) {
  cat("Absolute Income group counts for Continent:", cont, "\n")
  print(table(df_vars$`Income group`[df_vars$Continent == cont]))
  cat("\n")
}

for (cont in unique(df_vars$Continent)) {
  cat("Relative Income group distribution for Continent:", cont, "\n")
  print(prop.table(table(df_vars$`Income group`[df_vars$Continent == cont])))
  cat("\n")
}
```
Europe has the highest realtive amount of high income countries, followed by Oceania and America. The lowest is Africa.


- Using the distribution of income status across continents, identify which countries are the only ones in their income group across the continent. Discuss briefly.
```{r}
na.omit(df_vars$country[df_vars$Continent == 'Africa' & df_vars$`Income group` == 'High income'])
```
Since in the absolute Table only Africa has a 1 in any position, we could easly filtet for the high income country in africa
## f. Further Summary Statistics and Insights

- Create a table of average (mean and median) values for expenditure, youth unemployment rate and net migration rate separated into income status. Make sure that in the output, the ordering of the income classes is proper (i.e., L, LM, UM, H or the other way around). Briefly comment the results and any differences between the mean and median. 
```{r}
df_vars %>%
  mutate(`Income group` = factor(`Income group`, 
                                 levels = c("High income", "Upper middle income", 
                                            "Lower middle income", "Low income"))) %>%
  group_by(`Income group`) %>%
  summarise(
    avg_youth_unemp = mean(youth_unemployment, na.rm = TRUE),
    avg_expend = mean(education_expenditure, na.rm = TRUE),
    avg_mig_rate = mean(net_migration, na.rm = TRUE),
    median_youth_unemp = median(youth_unemployment, na.rm = TRUE),
    median_expend = median(education_expenditure, na.rm = TRUE),
    median_mig_rate = median(net_migration, na.rm = TRUE)
  )
```


- Look at the standard deviation and the interquartile range of the variables per income status instead of the location statistics above. Do you gain additional insights? Briefly comment the results. 
```{r}
df_vars %>%
  mutate(`Income group` = factor(`Income group`, 
                                 levels = c("High income", "Upper middle income", 
                                            "Lower middle income", "Low income"))) %>%
  group_by(`Income group`) %>%
  summarise(
    sd_youth_unemp = sd(youth_unemployment, na.rm = TRUE),
    sd_expend = sd(education_expenditure, na.rm = TRUE),
    sd_mig_rate = sd(net_migration, na.rm = TRUE),
    IQR_youth_unemp = IQR(youth_unemployment, na.rm = TRUE),
    IQR_expend = IQR(education_expenditure, na.rm = TRUE),
    IQR_mig_rate = IQR(net_migration, na.rm = TRUE)
  )
```
The results show that high-income countries have the most consistent patterns in youth unemployment and education expenditure, reflecting stable systems and lower variability. Lower-middle income countries display the highest volatility in net migration, suggesting greater economic or political instability. While low-income countries show less variability overall, this may reflect limited resources or structural constraints rather than true consistency. The NA group stands out with unusually high variation, indicating data quality issues or unclassified outliers that should be treated separately.

- Extend the analysis of the statistics median and IQR  to **each income status and continent combination**. Play around with displaying the resulting table. Use `pivot_longer()` and/or 
`pivot_wider()` to generate different outputs. 
Discuss the results as well as the readability of the different tables.
```{r}
df_vars %>%
  mutate(`Income group` = factor(`Income group`, 
                                 levels = c("High income", "Upper middle income", 
                                            "Lower middle income", "Low income"))) %>%
  group_by(`Income group`,`Continent`) %>%
  summarise(
    sd_youth_unemp = sd(youth_unemployment, na.rm = TRUE),
    sd_expend = sd(education_expenditure, na.rm = TRUE),
    sd_mig_rate = sd(net_migration, na.rm = TRUE),
    IQR_youth_unemp = IQR(youth_unemployment, na.rm = TRUE),
    IQR_expend = IQR(education_expenditure, na.rm = TRUE),
    IQR_mig_rate = IQR(net_migration, na.rm = TRUE)
  )

df_vars %>%
  mutate(`Income group` = factor(`Income group`, 
                                 levels = c("High income", "Upper middle income", 
                                            "Lower middle income", "Low income"))) %>%
  group_by(`Income group`, Continent) %>%
  summarise(
    sd_youth_unemp = sd(youth_unemployment, na.rm = TRUE),
    sd_expend = sd(education_expenditure, na.rm = TRUE),
    sd_mig_rate = sd(net_migration, na.rm = TRUE),
    IQR_youth_unemp = IQR(youth_unemployment, na.rm = TRUE),
    IQR_expend = IQR(education_expenditure, na.rm = TRUE),
    IQR_mig_rate = IQR(net_migration, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = Continent, values_from = c(sd_youth_unemp, sd_expend, sd_mig_rate,
                                                      IQR_youth_unemp, IQR_expend, IQR_mig_rate))
```
 All in all, the table with the continents in the header is more readable for comparing within and between income groups and continents. The drwaback is that the table becomes somewhat to long to read clearly on one screen, without scroling. Otherwise it is apparent from the results that oceania has the highest youth unemployment rat, as well as the lowest expenditure.
 
- Identify countries performing well in terms of both **youth unemployment** and **net migration rate** (top 25% in net migration and bottom 25% in youth unemployment within their continent).
```{r}
# Create a list to store the results
best_countries <- list()

# Loop through each unique continent
for (cont in unique(df_vars$Continent)) {
  
  # Subset data for this continent
  sub_df <- df_vars[df_vars$Continent == cont, ]
  
  # Define thresholds: low youth unemployment and high net migration
  youth_unemp_threshold <- quantile(sub_df$youth_unemployment, 0.25, na.rm = TRUE)
  net_mig_threshold <- quantile(sub_df$net_migration, 0.75, na.rm = TRUE)
  
  # Filter for countries meeting both criteria
  good_performers <- na.omit(sub_df$country[
    sub_df$youth_unemployment <= youth_unemp_threshold &
    sub_df$net_migration >= net_mig_threshold
  ])
  
  # Store in the list
  best_countries[[cont]] <- good_performers
}

# View result
best_countries
```

## g. Conditional Probabilities

Estimate the following based on the observed frequencies in the data:

- What is the (posterior or conditional) probability that a European country belongs to the high income group? What is the prior probability that a country belongs to the high income group?
```{r}
# P(European country)
p_european_country <- sum(df_vars$Continent == "Europe", na.rm = TRUE) / nrow(df_vars)

# P(High income and European)
p_high_and_europe <- sum(df_vars$Continent == "Europe" & df_vars$`Income group` == "High income", na.rm = TRUE) / nrow(df_vars)

# Posterior: P(High income | Europe)
p_high_income_given_europe <- p_high_and_europe / p_european_country

p_high_income_given_europe
```
Given a country is european, the probability it belongs to a high income group is 84%

- Given a country has high youth unemployment (above %25), what is the probability that it also has negative net migration?

```{r}
# P(Youth unemployment > 25%)
p_high_youth_unemp <- sum(df_vars$youth_unemployment > 25, na.rm = TRUE) / nrow(df_vars)

# P(Youth unemployment > 25% AND net migration < 0)
p_youth_unemp_and_neg_mig <- sum(df_vars$youth_unemployment > 25 & df_vars$net_migration < 0, na.rm = TRUE) / nrow(df_vars)

# P(Negative migration | High youth unemployment)
p_neg_mig_given_high_youth_unemp <- p_youth_unemp_and_neg_mig / p_high_youth_unemp

p_neg_mig_given_high_youth_unemp
```
The probability of a high youth unemployment the probabilty is that it has negative net migration is 61%
## h. Simpson’s Paradox Analysis

Investigate whether an overall trend in youth unemployment rate in the high and low income groups reverses when analyzed at the continent level. E.g., does the youth unemployment rate appear lower in low-income countries overall, but higher when controlling for continent? Explain the results and possible reasons behind this paradox.

To explore whether Simpson’s Paradox occurs in this dataset, we analyze youth unemployment rates across income groups at two levels:

1. **Overall (global) level**
2. **Within each continent**

This allows us to check whether the observed global trend reverses when controlling for continent.

---

### 1. Overall Youth Unemployment by Income Group

```{r simpson-overall-plot}
ggplot(df_vars, aes(x = `Income group`, y = youth_unemployment, fill = `Income group`)) +
  geom_boxplot() +
  theme_minimal() +
  labs(
    title = "Youth Unemployment by Income Group (Overall)",
    x = "Income Group",
    y = "Youth Unemployment Rate (%)"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
The overall boxplot suggests that, on average, youth unemployment appears lower in low-income countries compared to higher-income groups.

### 2. Youth Unemployment by continent
```{r simpson-by-continent-plot}
ggplot(df_vars, aes(x = `Income group`, y = youth_unemployment, fill = `Income group`)) +
  geom_boxplot() +
  facet_wrap(~ Continent) +
  theme_minimal() +
  labs(
    title = "Youth Unemployment by Income Group Within Continents",
    x = "Income Group",
    y = "Youth Unemployment Rate (%)"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The within-continent trends differ from the global pattern. In regions such as Africa and Asia, low-income countries tend to show **higher youth unemployment** than high-income countries, **reversing** the global trend.

This indicates a potential case of **Simpson’s Paradox**: while the aggregated data suggests that low-income countries have lower unemployment, this is due to their concentration in regions with generally lower overall unemployment. When controlling for continent, the underlying association between income level and unemployment becomes visible, revealing that lower income is often linked to higher youth unemployment within regions.

Thus, this analysis provides evidence for Simpson’s Paradox in the relationship between income group and youth unemployment.


## i. Data Export

Export the final tidy dataset from e. as a **CSV** with:`;` as a separator; `.` representing missing values; no row names included.  Upload the `.csv` to TUWEL, together with the submission.